---
title: "Alternative chart types practice"
author: "Aakriti Poudel"
date: 2026-03-04
toc: true
---

::: {.callout-note}
This template accompanies the [alternative chart types practice slides](https://eds-240-data-viz.github.io/course-materials/lecture-slides/SLIDES-alt-charts.html#/title-slide){target="_blank}. Please be sure to cross-reference the slides, which contain important information and additional context!
:::

# Waffle Chart (Sasquatch sightings)

- Typefaces used: Ultra & Josefin Sans
- Colors used: "#357266", "#FFB813", "#983A06", "#005F71", "#757473", "#EFEFEF"

```{r}
#| eval: true
#| echo: true
#| message: false
#| warning: false
#| fig-align: "center"
##~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
##                                    setup                                 ----
##~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~

#..........................load packages.........................
library(tidyverse)
library(waffle)
library(showtext)

#..........................import data...........................
bigfoot <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2022/2022-09-13/bigfoot.csv')
```


```{r}
# ============================================================
# Bigfoot Sightings Waffle Chart — California
# Data: TidyTuesday 2022-09-13 | BFRO
# ============================================================

library(tidyverse)
library(waffle)
library(showtext)


# ── 1. LOAD FONTS ────────────────────────────────────────────
font_add_google("Lato", "lato")
font_add_google("Playfair Display", "playfair")
showtext_auto()


# ── 2. READ DATA ─────────────────────────────────────────────
bigfoot <- readr::read_csv(
  "https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2022/2022-09-13/bigfoot.csv"
)


# ── 3. INSPECT ───────────────────────────────────────────────
glimpse(bigfoot)
# Key columns we'll use:
#   state     — U.S. state of the sighting
#   season    — reported season (Spring / Summer / Fall / Winter / Unknown)
#   date      — date of sighting (character → needs parsing if used)


# ── 4. DATA CLEANING & WRANGLING ─────────────────────────────

bf_clean <- bigfoot |>

  # 4a. Keep only California sightings
  filter(state == "California") |>

  # 4b. Drop rows with missing or "Unknown" season
  filter(!is.na(season), season != "Unknown") |>

  # 4c. Enforce factor order (Spring → Summer → Fall → Winter)
  mutate(
    season = factor(
      season,
      levels = c("Spring", "Summer", "Fall", "Winter")
    )
  ) |>

  # 4d. Select only the columns we need
  select(season, state, date, latitude, longitude)

# Quick sanity check
bf_clean |>
  count(season, name = "n_sightings") |>
  mutate(pct = round(n_sightings / sum(n_sightings) * 100, 1))


# ── 5. SUMMARISE → NAMED VECTOR FOR waffle() ─────────────────
# waffle() expects a named numeric vector; each unit = 1 square

season_counts <- bf_clean |>
  count(season) |>
  deframe()                    # tibble col1 → names, col2 → values

# Result looks like:
#  Spring Summer   Fall Winter
#      81    134     93     28

# Scale down if total is large (each square = k sightings)
# Here total ≈ 336; keep 1 square = 1 sighting for fidelity
# Uncomment below to use 1 square = 5 sightings:
# season_counts <- round(season_counts / 5)


# ── 6. COLOUR PALETTE ────────────────────────────────────────
season_colors <- c(
  Spring = "#2d6a4f",   # forest green
  Summer = "#f4a61d",   # golden yellow
  Fall   = "#7b3a10",   # burnt sienna
  Winter = "#2176ae"    # steel blue
)


# ── 7. PLOT ───────────────────────────────────────────────────
BG <- "#ffffff" 

p <- waffle(
  parts      = season_counts,
  rows       = 17,                # number of rows in the grid
  colors     = season_colors,
  size       = 0.4,               # line size between squares (tile border width)
  flip       = FALSE,
  legend_pos = "bottom"
) +

  # This controls the gap color between squares
  scale_color_manual(values = rep(BG, length(season_counts)), guide = "none") +

  # Remove the default ggplot padding around the data range.
  # Without this, ggplot adds ~5% whitespace beyond the last tile on each axis,
  # which shows up as a white/off-color strip at the panel edges.
  scale_x_continuous(expand = c(0, 0)) +
  scale_y_continuous(expand = c(0, 0)) +

  labs(
    title    = "Summer is the season of Bigfoot sightings in CA",
    subtitle = "Winter, on the other hand, is a rare time to spot Sasquatch",
    caption  = "Source: Bigfoot Field Researchers Organization (BFRO)"
  ) +

  theme_minimal(base_family = "lato") +

  theme(
    # Titles
    plot.title    = element_text(
      family    = "playfair",
      face      = "bold",
      size      = 16,
      color     = "#1a1a2e",
      margin    = margin(b = 4)
    ),
    plot.subtitle = element_text(
      family = "lato",
      size   = 11,
      color  = "#444444",
      margin = margin(b = 12)
    ),
    plot.caption  = element_text(
      size   = 8,
      color  = "#888888",
      hjust  = 1,
      margin = margin(t = 10)
    ),

    # Legend
    legend.position  = "bottom",
    legend.title     = element_blank(),
    legend.text      = element_text(size = 10, family = "lato"),
    legend.key.size  = unit(12, "pt"),
    legend.spacing.x = unit(8, "pt"),

    # Background
    # NOTE: waffle uses geom_tile internally, so panel.background alone
    # has no effect. Use theme_enhance_waffle() below to strip the default
    # panel, then set plot.background to fill the entire canvas.
    plot.background  = element_rect(fill = BG, color = NA),
    panel.background = element_rect(fill = BG, color = NA),
    panel.border     = element_blank(),
    panel.grid       = element_blank(),
    axis.text        = element_blank(),
    axis.ticks       = element_blank(),

    # Margins
    plot.margin = margin(20, 24, 14, 24)
  )

# theme_enhance_waffle() removes the internal grey panel that waffle draws
# by default — must come AFTER your theme() call or it gets overridden
p <- p + theme_enhance_waffle()

print(p)


# ── 8. SAVE ───────────────────────────────────────────────────
ggsave(
  filename = "bigfoot_waffle_CA.png",
  plot     = p,
  width    = 8,
  height   = 5.5,
  dpi      = 300,
  bg       = BG
)
message("✓ Plot saved → bigfoot_waffle_CA.png")
```


# Bump chart (highest paying occupations)

- Typefaces used: Passion One & Oxygen
- Colors used: `viridisLite::magma()` & "#2A114E", "#6D6B71", "#FFF8F4"

```{r}
##~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
##                                    setup                                 ----
##~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~

#..........................load packages.........................
library(tidyverse)
library(ggbump)
library(ggtext)
library(showtext)

#..........................import data...........................
jobs <- read_csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2019/2019-03-05/jobs_gender.csv")
```

# Stream graph (energy production)

- Typefaces used: Exo 2 & Open Sans
- Colors used: `paletteer::paletteer_d("MetBrewer::Archambault", n = 7)`

```{r}
##~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
##                                    setup                                 ----
##~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~

#..........................load packages.........................
library(tidyverse)
library(ggstream)
library(ggtext)
library(showtext)

#..........................import data...........................
capacity <- readr::read_csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2022/2022-05-03/capacity.csv")
```











